### 2021-06-18 papers/thesis/submitted/acknowledgements/acknowledgements.tex
Submitted version saved in new dir

I would like to thank the MPE CDT staff for any knowledge and wisdom they have bestowed, and in particular say a huge thank you to Colin Cotter for his guidance, understanding and general enthusiasm (it meant a lot). And to all my fellow \enquote{Charlies} -- I truly mean it when I say you are all such great, smart and kind people, and you all deserve the absolute best for the future.

I thank my family and friends for all their support and words of encouragement -- it all helped a lot.

A special thanks goes to my partner Jess for putting up with me and helping me when things got tough during my 4 years at Imperial College London.

Finally, to Dr. Sheehan Olver -- I cannot thank you enough for all your help, wisdom, advice, kindness, understanding and patience you have given me. I may not have expressed my gratitude enough over the past few years, but you have truly been a great supervisor to me.

>>>>

I would like to thank the MPE CDT staff for the knowledge and wisdom they have bestowed, and in particular say a huge thank you to Colin Cotter for his guidance, assistance and general enthusiasm (it meant a lot). And to all my fellow \enquote{Charlies} -- I truly mean it when I say you are all such great, smart and kind people, and you all deserve the absolute best for the future.

I thank my friends for their support and the fun times they have provided me with over these few years. I also thank my family for all their support and encouragement, not just during my PhD but throughout my life as well to help me get to this point.

A special thanks goes to my wonderful partner Jess for putting up with me and helping me when things got tough during my time at Imperial College London.

Finally, to Dr. Sheehan Olver -- I cannot thank you enough for all your help, wisdom, advice, kindness, understanding and patience you have given me. I may not have expressed my gratitude enough, but you have truly been a great supervisor to me.

### 2021-06-18 papers/thesis/submitted/ch4-spherical-caps/spherical-caps-old.tex
Submitted version saved in new dir

\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
%\geometry{landscape}                		% Activate for rotated page geometry
\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex

\graphicspath{ {images/} }

\include{bsthesisnewcommands}
\include{bsmacros}
\include{somacros}

\title{Sparse spectral methods for partial differential equations on spherical caps}
\author{Ben Snowball, Sheehan Olver\thanks{Department of Mathematics, Imperial College, London, England}}
%\date{}							% Activate to display a given date or no date

\begin{document}

\maketitle

\begin{abstract}
In recent years, sparse spectral methods for solving partial differential equations have been derived using hierarchies of classical orthogonal polynomials on intervals, disks, disk-slices and triangles. In this work we extend the methodology to a hierarchy of non-classical multivariate orthogonal polynomials on spherical caps. The entries of discretisations of partial differential operators can be effectively computed using formulae in terms of (non-classical) univariate orthogonal polynomials. We demonstrate the results on partial differential equations involving the spherical Laplacian and biharmonic operators, showing spectral convergence. 
\end{abstract}

@@

For a function $f$, with coefficients vector $\bm{f}$ for expansion in the $\{\scopnki\}$ basis as determined via the method in Section \ref{subsection:expandingfunctions} up to order $N$, we can use the Clenshaw algorithm to evaluate the function at a point $(x,y,z) \in \Omega$ as follows. Let $A_n, B_n, \Dnt, C_n$ be the Clenshaw matrices from Definition \ref{def:clenshawmats}, and define the rearranged coefficients vector $\tilde{\bm{f}}$ via
\begin{align*}
	{\bm{f}}_n &:= 
		\begin{pmatrix}
			f_{n,0,0} \\
			f_{n,1,0} \\
			f_{n,1,1} \\
			\vdots \\
			f_{n,n,0} \\
			f_{n,n,1}
		\end{pmatrix} \in \R^{2(N+1)} \quad \text{for } n = 1,2,\dots,N, \quad
	{\bm{f}}_0 = f_{0,0,0} \in \R, \\
	{\bm{f}} &:= 
		\begin{pmatrix}
			{\bm{f}}_0 \\
			\vdots \\
			{\bm{f}}_N
		\end{pmatrix} \in \R^{(N+1)^2}.
\end{align*}
The trivariate Clenshaw algorithm works similar to the bivariate Clenshaw algorithm introduced in \cite{olver2019triangle} for expansions in the triangle:
\begin{align*}
	\quad &\text{1) } \text{Set } \bm{\xi}_{N+2} = \bold{0}, \: \bm{\xi}_{N+2} = \bold{0}. \\
	\quad &\text{2) } \text{For } n = N:-1:0 \\
	\quad & \quad \quad \quad \text{set } \bm{\xi}_{n}^T = {{\bm{f}}_n}^T - \bm{\xi}_{n+1}^T D^T_n (B_n - G_n(x,y,z)) -  \bm{\xi}_{n+2}^T D^T_{n+1}C_{n+1} \\
	\quad &\text{3) } \text{Output: } f(x,y,z) \approx \bm{\xi}_{0} \: \bigscopta_{0}  = \xi_{0} \: \scopa_0
\end{align*}

@@

\begin{figure}[t]
	\centering % <-- added
	\includegraphics[scale=0.35]{complexity-Nend=1000}
	\caption{Time in seconds to build and solve the system $\big[\DeltaS + v(x,y,z) \big] \: u(x,y,z) = f(x,y,z)$, for a rotationally invariant $v(x,y,z) = v(z)$. This demonstrates that the approach is roughly of order $\bigO(N^2)$, where $N$ is the degree to which we approximate the solution. Here, we used $f = -2e^{x}yz(2+x) + (z - \alpha) e^{x} (y^3 + z^2 y - 4xy - 2y)$ and $v(x,y,z) = v(z) = \cos(z)$.}
	\centering
	\label{fig:complexity}
\end{figure}

@@

In Figure~\ref{fig:complexity} we plot the time taken\footnote{measured using the \enquote{@belapsed} macro from the BenchmarkTools.jl package \cite{BenchmarkTools.jl-2016} in Julia.} to construct the operator for $\DeltaS + v(x,y,z)$, with a rotationally invariant $v(x,y,z) = v(z) = \cos z$, and solve a zero boundary condition Helmholtz problem. The plot demonstrates that as we increase the degree of approximation $N$, we achieve a complexity of an optimal $\bigO(N^2)$.

What about other boundary conditions? One simple extension is the case where the value on the boundary takes that of a function depending only on $x$ and $y$, i.e. $c = c(x,y)$. In this case, the problem
\begin{align*}
	\begin{cases}
    		\DeltaS u(x,y,z) + k^2 \: v(x,y,z) \; u(x,y,z) = f(x,y,z) &\quad \text{in } \Omega \\
		u(x,y,z) = c(x,y) &\quad \text{on } \partial \Omega
	\end{cases}
\end{align*}
is equivalent to letting $u(x,y,z) = \tilde{u}(x,y,z)+ c(x,y)$ and solving
\begin{align*}
	\begin{cases}
    		\DeltaS \tilde{u}(x,y,z) + k^2 \: v(x,y,z) \; \tilde{u}(x,y,z) = f(x,y,z) - k^2 \: v(x,y,z) \: c(x,y) - \DeltaS \: c(x,y)  \quad \text{in } \Omega \\
		\tilde{u}(x,y,z) = 0 \quad \text{on } \partial \Omega
	\end{cases}
\end{align*}
for $\tilde u$. This new problem is then a zero boundary condition Helmholtz problem with right hand side 
\begin{align*}
	g(x,y,z) &:= f(x,y,z) - k^2 \: v(x,y,z) \: c(x,y) - \DeltaS \: c(x,y)
\end{align*}
for $(x,y,z) \in \Omega$. Notice that the spherical Laplacian applied to $c(x,y)$, expanded in the $\bigscopN^{(1)}$ basis with coefficients vector $\vec c = (c_{n,k,i})$, is just
\begin{align*}
	\DeltaS \: c(x,y) = {1 \over \rho(z)^2} \sum_{n=0}^N \sum_{i=0}^1 c_{n,n,i} \: \pmpxm{\theta}{2} \ch_{n,i}(\theta) = - {1 \over \rho(z)^2} \sum_{n=0}^N \sum_{i=0}^1 n^2 \: c_{n,n,i} \: \ch_{n,i}(\theta)
\end{align*}
since the coefficients $\{c_{n,k,i}\}$ for such a function are zero for $k < n$ due to the dependence on $x$ and $y$ only, which are precisely the Fourier coefficients of $c(\cos \theta, \sin \theta)$. Thus, since the function $c(x,y)$ is known, it is simple to evaluate $\pmpxm{\theta}{2} c(x,y)$ and hence one can obtain the coefficients for the expansion of $g(x,y,z)$ in the $\bigscopN^{(1)}$ basis in the usual manor.

@@

Our last erxample is the biharmonic equation: find $u(x,y,z)$ given a function $f(x,y,z)$ such that:
\begin{align}
	\begin{cases}
    		\DeltaS^2 u(x,y,z) = f(x,y,z) &\quad \text{in } \Omega \\
		u(x,y,z) = 0, \quad \frac{\partial u}{\partial n}(x,y,z) = \grad_S \: u(x,y,z) \cdot \unitvec{n}(x,y,z) = 0 &\quad \text{on } \partial \Omega
	\end{cases}
	\label{eqn:biharmonic}
\end{align}
where $\DeltaS^2$ is the Biharmonic operator, noting the imposition of zero Dirichlet and Neumann boundary conditions on $u$. For clarity, we reiterate that the unit normal vector in this sense is simply $\unitvec{n}(x,y,z) = \unitvec{n}(\vec x) := {\vec x \over \norm{\vec x}} = \vec x$ (see Section \ref{subsection:furtherdiffoperators}). In Figure \ref{fig:biharmonic} we see the solution to the Biharmonic equation (\ref{eqn:biharmonic}) in the spherical cap $\Omega$. In Figure \ref{fig:biharmonic} we also show the norms of each block of calculated coefficients of the approximation for four more complex right-hand sides of the biharmonic equation with $N = 200$, that is, $(N+1)^2 = 40,401$ unknowns. Once again, the rate of decay in the coefficients is a proxy for the rate of convergence of the computed solution, and we see that we achieve exponential convergence for these more complex functions.

%
\section{Conclusions}

We have shown that trivariate orthogonal polynomials can lead to sparse discretizations of general linear PDEs on spherical cap domains, with Dirichlet boundary conditions on the $z = \alpha \in (0,1)$ boundary. We have provided a detailed practical framework for the application of the methods described for quadratic surfaces of revolution \cite{olver2020orthogonal}, by utilising the non-classical 1D OPs on the interval $[\alpha, 1]$ with the weight $(z - \alpha)^a \: (1-z^2)^{b/2}$ defined for the disk-slice case \cite{snowball2019sparse}. Generalisation to spherical bands ($\alpha \leq z \leq \beta$) is straightforward.  This work thus forms a building block in developing an $hp-$finite element method to solve PDEs on the sphere by using spherical band and spherical cap shaped elements.

This work also serves as a stepping stone to constructing similar methods to solve partial differential equations on other 3D sub-domains of the sphere---it is clear from the construction in this paper that discretizations of spherical gradients and Laplacian's are sparse on other suitable sub-components of the sphere. The resulting sparsity in high-polynomial degree discretizations presents an attractive alternative to methods based on bijective mappings (e.g., \cite{DGShallowWater,FEMShallowWater,boyd2005sphere}). Constructing these sparse spectral methods for surface PDEs on  spherical triangles is future work, and has applications in weather prediction \cite{staniforth2012horizontal}, though it is not yet clear how to directly construct the necessary orthogonal polynomials.

The next stage is to develop an orthogonal basis for the tangent space of the spherical cap (or band), and obtain sparse differential operators for gradient, divergence etc. On the complete sphere, the vector spherical harmonics that form the orthogonal basis are simply the gradients and perpendicular gradients of the scalar spherical harmonics \cite{barrera1985vector} which has been used effectively for solving PDEs on the sphere \cite{vasil2019tensor,lecoanet2019tensor} -- however, we do not have that luxury for the spherical cap or band, and hence the choice of basis will not be as straightforward.

%\noindent{\bf Acknowledgements}:  We would like to thank the anonymous referees for their helpful comments.  The second author was supported in part by the Leverhulme Trust Research Project Grant RPG-2019-144 ``Constructive approximation theory on and inside algebraic curves and surfaces.
%%
%\appendix
%%
%\section{P-finite element methods using sparse operators}\label{Appendix:PFEM}
%\bstodoinline{Is this worth including?}
%\sotodoinline{It's already a very long paper so probably better to skip this. But include it in your thesis} \bstodoinline{Yes thought so too}
%%\begin{figure}[t]
%%	\begin{subfigure}{0.3\textwidth}
%%	\includegraphics[scale=0.28]{solution-pfem-poisson-diskslice-alpha=0p2-beta=0p8-f=Wycosx}
%%        	%\label{fig:solutionblocknorms-poisson}
%%	\centering
%%	\end{subfigure}
%%	\begin{subfigure}{0.3\textwidth}
%%	\includegraphics[scale=0.3]{solution-poisson}
%%	\centering
%%	%\label{fig:solution-poisson}
%%	\end{subfigure}
%%	\begin{subfigure}{0.3\textwidth}
%%	\includegraphics[scale=0.3]{solution-trapezium-helmholtz-k=100-n=1500}
%%	%\label{fig:solution-poisson}
%%	\centering
%%	\end{subfigure}
%%	\centering
%%	\caption{Left: The computed solution to $\Delta u = f$ with zero boundary conditions with $f(x,y) = \Wiii(x,y) y \cos(x)$ in the disk-slice using the $p$-FEM approach with a single element. Centre: The computed solution to $\Delta u = f$ with zero boundary conditions with $f(x,y) = 1 + \text{erf}(5(1 - 10((x - 0.5)^2 + y^2)))$ in the half-disk. Right: The computed solution to $\Delta u + k^2 \: v \: u = f$ with zero boundary conditions with $f(x,y) = (1-x) \: x \: y \: (1- \half x - y) \: e^x$, $v(x,y) = 1 - (3(x-1)^2 + 5y^2)$ and $k = 100$. in the trapezium.}
%%	\centering
%%	\label{fig:appendixfigs}
%%\end{figure}
%We follow the method of \cite{beuchler2006new} to construct a sparse $p$-finite element method in terms of the operators constructed above, with the benefit of ensuring that the resulting discretisation is symmetric. Consider the 3D Dirichlet problem on a domain $\Omega$:
%	\begin{cases}
%         - \rho(z)^2 \: \DeltaS u(x,y,z) = \rho(z)^2 \: f(x,y,z) \quad \text{in } \Omega \\
%         u = 0 \quad \text{on } \partial \Omega
%         \end{cases}
%This has the weak formulation
%	L(v) := \int_\Omega \rho^2 \: f \: v \: \D \sigma(\bm{x}, z) = \int_\Omega \big(\rho \: \nabla u\big) \cdot \big(\rho \: \nabla v\big) \: \D \sigma(\bm{x}, z) =: a(u,v)
% for any test function $v \in V := H_0^1(\Omega) = \{v \in H^1(\Omega) \quad | \quad v \rvert_{\partial \Omega} = 0 \}$.
%In general, we would split a sphere domain into spherical band elements with spherical caps over the poles, and use a $p$-finite element method to solve PDEs on the sphere using these elements $\{\element_n\}$. While spherical bands are an extension to this work, we note that we can still apply a $p$-FEM using two half-spheres ($\alpha = 0$). In this section, we limit our discretisation to a single element, that is we let $\element = \Omega$ for a spherical cap domain. We can choose our finite dimensional space $V_p = \{v_p \in V \quad | \quad {\rm deg}\,(v_p \rvert_\element) \le p\}$ for some $p \in \N$.
%We seek $u_p \in V_p$ s.t.
%\begin{align}
%	L(v_p) = a(u_p,v_p) \quad \forall \: v_p \in V_p.
%	\label{eqn:FEMweakform}
%\end{align}
%Define 
%	\Lambda^{(a)} :=  \ip< \bigscopa_p, \: {\bigscopa_p}^\top >_{\scopa} := \int_\element \bigscopa_p \: {\bigscopa_p}^\top \: \genjacw^{(a,0)} \: \D \sigma(\bm{x}, z) = \int_\element \bigW_p^{(a)} \: {\bigscopa_p}^\top \: \D \sigma(\bm{x}, z)
%for any $a$. Note that due to orthogonality this is a diagonal matrix. By choosing suitable bases for our FE spaces, we can rewrite (\ref{eqn:FEMweakform}) in matrix-vector form. We can choose a basis for $V_p$ by using the weighted orthogonal polynomials on $\element$ with parameter $a = 1$, the $\bigW_p^{(1)}$ OPs, and we can expand the function $f$ in the $\bigscopi_p$ basis. 
%Let $\mathbf{u}, \mathbf{v}$ be the coefficient vectors of the expansions of $u_p, v_p \in V_p$ respectively in the $V_p$ basis ($\bigWiii$ OPs). Then,
%	a(u_p,v_p) &= \int_\element \big(\rho \: \nabla u\big) \cdot \big(\rho \: \nabla v\big) \: \D \sigma(\bm{x}, z) \\
%	&= \int_\element \Big( \rhoppphi u_p \: \rhoppphi v_p + \pptheta u_p \: \pptheta v_p \Big) \: \D \sigma(\bm{x}, z) \\
%	&= \int_\element \Big( \mathbf{v}^\top {W_\varphi^{(1)}}^\top \bigscopo_p {\bigscopo_p}^\top W_\varphi^{(1)} \mathbf{u} \nonumber \\
%					& \quad \quad \quad + \mathbf{v}^\top ({T_W^{(1)\to(0)} D_\theta})^\top \bigscopa_p {\bigscopa_p}^\top T_W^{(1)\to(0)} D_\theta \mathbf{u}  \Big) \: \D \sigma(\bm{x}, z) \\
%	&= \mathbf{v}^\top \: \Big( {W_\varphi^{(1)}}^\top \Lambda^{(0)}  W_\varphi^{(1)} + ({T_W^{(1)\to(0)} D_\theta})^\top \Lambda^{(0)} T_W^{(1)\to(0)} D_\theta \Big) \: \mathbf{u}.
%Further, let $\mathbf{f}$ is the coefficient vector for the expansion of the function $f(x,y,z)$ in the $\bigscopi_p$ OP basis and let $P$ be the operator for multiplication by $\rho(z)^2$, that is the operator matrix defined by $P^\top \bigscopi_p(x,y,z) = \rho(z)^2 \: \bigscopi_p(x,y,z)$. Then,
%	L(v_p) &= \int_\element \: v_p \: \rho^2 \: f \: \D \sigma(\bm{x}, z) \\
%	&= \int_\element \: \mathbf{v}^\top \: \bigW_p^{(1)} \: {\bigscopi_p}^\top \: P \: \mathbf{f} \: \D \sigma(\bm{x}, z) \\	
%	&= \mathbf{v}^\top \Lambda^{(1)} \: P \: \mathbf{f}.
%Since (\ref{eqn:FEMweakform}) is equivalent to stating that
%	L(\genjacw^{(1)} \scopnkii) = a(u_p, \genjacw^{(1)} \scopnkii) \quad \forall \: n = 0,\dots,p, \: k = 0,\dots,n, \: i = 0,1
%(i.e. holds for all basis functions of $V_p$) by choosing $v_p$ as each basis function, we can equivalently write the linear system for our finite element problem as:
%	A\mathbf{u} = \tilde{\mathbf{f}}.
%where the (element) stiffness matrix $A$ is defined by 
%	A =  {W_\varphi^{(1)}}^\top \Lambda^{(0)}  W_\varphi^{(1)} + ({T_W^{(1)\to(0)} D_\theta})^\top \Lambda^{(0)} T_W^{(1)\to(0)} D_\theta
%and the load vector $\tilde{\mathbf{f}}$ is given by 
%	\tilde{\mathbf{f}} = \Lambda^{(1)} \: P \: \mathbf{f}.
%Note that since we have sparse operator matrices for partial derivatives and basis-transform, we obtain a symmetric sparse (element) stiffness matrix, as well as a sparse operator matrix for calculating the load vector (rhs).

\bibliography{spherical-caps}

\end{document}

>>>>

\chapter{Spherical Caps}\label{CH:SCs}

While the work in the previous chapter looked at developing a sparse spectral method inside a two dimensional domain, we move on to investigating the realm of a surface in three dimensional space. Specifically, we look to extend the methodology to a hierarchy of non-classical multivariate orthogonal polynomials on spherical caps, which we will formally define. The entries of discretisations of partial differential operators can be effectively computed using formulae in terms of (non-classical) univariate orthogonal polynomials. We demonstrate the results on partial differential equations involving the spherical Laplacian and Biharmonic operators, showing spectral convergence.

@@

For a function $f$, with coefficients vector $\bm{f}$ for expansion in the $\{\scopnki\}$ basis as determined via the method in Section \ref{subsection:expandingfunctions} up to order $N$, we can use the Clenshaw algorithm to evaluate the function at a point $(x,y,z) \in \Omega$ as follows. Let $A_n, B_n, \Dnt, C_n$ be the Clenshaw matrices from Definition \ref{def:clenshawmats}, and define the rearranged coefficients vector $\tilde{\bm{f}}$ via
\begin{align*}
	\tilde{\bm{f}}_n &:= 
		\begin{pmatrix}
			f_{n,0,0} \\
			f_{n,1,0} \\
			f_{n,1,1} \\
			\vdots \\
			f_{n,n,0} \\
			f_{n,n,1}
		\end{pmatrix} \in \R^{2(N+1)} \quad \text{for } n = 1,2,\dots,N, \quad
	\tilde{\bm{f}}_0 = f_{0,0,0} \in \R, \\
	\tilde{\bm{f}} &:= 
		\begin{pmatrix}
			\tilde{\bm{f}}_0 \\
			\vdots \\
			\tilde{\bm{f}}_N
		\end{pmatrix} \in \R^{(N+1)^2}.
\end{align*}
The trivariate Clenshaw algorithm works similar to the bivariate Clenshaw algorithm introduced in \cite{olver2019triangle} for expansions in the triangle:
\begin{align*}
	\quad &\text{1) } \text{Set } \bm{\xi}_{N+2} = \bold{0}, \: \bm{\xi}_{N+2} = \bold{0}. \\
	\quad &\text{2) } \text{For } n = N:-1:0 \\
	\quad & \quad \quad \quad \text{set } \bm{\xi}_{n}^T = {\tilde{\bm{f}}_n}^T - \bm{\xi}_{n+1}^T D^T_n (B_n - G_n(x,y,z)) -  \bm{\xi}_{n+2}^T D^T_{n+1}C_{n+1} \\
	\quad &\text{3) } \text{Output: } f(x,y,z) \approx \bm{\xi}_{0} \: \bigscopta_{0}  = \xi_{0} \: \scopa_0
\end{align*}

@@

\begin{figure}[t]
	\centering % <-- added
	\includegraphics[scale=0.35]{complexity-Nend=1000}
	\caption{Complexity plot showing the length of time taken to construct the operator matrix for $\DeltaS + v(x,y,z)$, with rotationally invariant potential $v(x,y,z) = \cos z$. This shows that the time to build and solve the system $\big[\DeltaS + v(x,y,z) \big] \: u(x,y,z) = f(x,y,z)$ is roughly of order $\bigO(N^2)$, where $N$ is the degree to which we approximate the solution. Here, we used $f = -2e^{x}yz(2+x) + (z - \alpha) e^{x} (y^3 + z^2 y - 4xy - 2y)$.}
	\centering
	\label{fig:complexity}
\end{figure}

@@

In Figure~\ref{fig:complexity} we plot the time taken\footnote{measured using the \enquote{@belapsed} macro from the BenchmarkTools.jl package \cite{BenchmarkTools.jl-2016} in Julia.}  to construct the operator for $\DeltaS + v(x,y,z)$, this time with a rotationally invariant $v(x,y,z) = \cos z$, and solve a zero boundary condition Helmholtz problem. The plot demonstrates that as we increase the degree of approximation $N$, we achieve a complexity of an optimal $\bigO(N^2)$.

What about other boundary conditions? One simple extension is the case where the value on the boundary takes that of a function depending only on $x$ and $y$, i.e. $c = c(x,y)$. In this case, the problem
\begin{align*}
	\begin{cases}
    		\DeltaS u(x,y,z) + k^2 \: v(x,y,z) \; u(x,y,z) = f(x,y,z) &\quad \text{in } \Omega \\
		u(x,y,z) = c(x,y) &\quad \text{on } \partial \Omega
	\end{cases}
\end{align*}
is equivalent to letting $u(x,y,z) = \tilde{u}(x,y,z)+ c(x,y)$ and solving
\begin{align*}
	\begin{cases}
    		\DeltaS \tilde{u}(x,y,z) + k^2 \: v(x,y,z) \; \tilde{u}(x,y,z) = f(x,y,z) - k^2 \: v(x,y,z) \: c(x,y) - \DeltaS \: c(x,y)  \quad \text{in } \Omega \\
		\tilde{u}(x,y,z) = 0 \quad \text{on } \partial \Omega
	\end{cases}
\end{align*}
for $\tilde u$. This new problem is then a zero boundary condition Helmholtz problem with right hand side 
\begin{align*}
	g(x,y,z) &:= f(x,y,z) - k^2 \: v(x,y,z) \: c(x,y) - \DeltaS \: c(x,y)
\end{align*}
for $(x,y,z) \in \Omega$. Notice that the spherical Laplacian applied to $c(x,y)$, expanded in the $\bigscopN^{(1)}$ basis with coefficients vector $\vec c = (c_{n,k,i})$, is just
\begin{align*}
	\DeltaS \: c(x,y) = {1 \over \rho(z)^2} \sum_{n=0}^N \sum_{i=0}^1 c_{n,n,i} \: \pmpxm{\theta}{2} \ch_{n,i}(\theta) = - {1 \over \rho(z)^2} \sum_{n=0}^N \sum_{i=0}^1 n^2 \: c_{n,n,i} \: \ch_{n,i}(\theta)
\end{align*}
since the coefficients $\{c_{n,k,i}\}$ for such a function are zero for $k < n$ due to the dependence on $x$ and $y$ only. Thus, since the function $c(x,y)$ is known, it is simple to evaluate $\pmpxm{\theta}{2} c(x,y)$ and hence one can obtain the coefficients for the expansion of $g(x,y,z)$ in the $\bigscopN^{(1)}$ basis in the usual manor.

@@

Find $u(x,y,z)$ given a function $f(x,y,z)$ such that:
\begin{align}
	\begin{cases}
    		\DeltaS^2 u(x,y,z) = f(x,y,z) &\quad \text{in } \Omega \\
		u(x,y,z) = 0, \quad \frac{\partial u}{\partial n}(x,y,z) = \grad_S \: u(x,y,z) \cdot \unitvec{n}(x,y,z) = 0 &\quad \text{on } \partial \Omega
	\end{cases}
	\label{eqn:biharmonic}
\end{align}
where $\DeltaS^2$ is the Biharmonic operator, noting the imposition of zero Dirichlet and Neumann boundary conditions on $u$. For clarity, we reiterate that the unit normal vector in this sense is simply $\unitvec{n}(x,y,z) = \unitvec{n}(\vec x) := {\vec x \over \norm{\vec x}} = \vec x$ (see Section \ref{subsection:furtherdiffoperators}). In Figure \ref{fig:biharmonic} we see the solution to the Biharmonic equation (\ref{eqn:biharmonic}) in the spherical cap $\Omega$. In Figure \ref{fig:biharmonic} we also show the norms of each block of calculated coefficients of the approximation for four more complex right-hand sides of the biharmonic equation with $N = 200$, that is, $(N+1)^2 = 40,401$ unknowns. Once again, the rate of decay in the coefficients is a proxy for the rate of convergence of the computed solution, and we see that we achieve algebraic convergence for these more complex functions.

\section{$p$-finite element methods using sparse operators}

%\begin{figure}[t]
%	\begin{subfigure}{0.3\textwidth}
%	\includegraphics[scale=0.28]{solution-pfem-poisson-diskslice-alpha=0p2-beta=0p8-f=Wycosx}
%        	%\label{fig:solutionblocknorms-poisson}
%	\centering
%	\end{subfigure}
%	\begin{subfigure}{0.3\textwidth}
%	\includegraphics[scale=0.3]{solution-poisson}
%	\centering
%	%\label{fig:solution-poisson}
%	\end{subfigure}
%	\begin{subfigure}{0.3\textwidth}
%	\includegraphics[scale=0.3]{solution-trapezium-helmholtz-k=100-n=1500}
%	%\label{fig:solution-poisson}
%	\centering
%	\end{subfigure}
%	\centering
%	\caption{Left: The computed solution to $\Delta u = f$ with zero boundary conditions with $f(x,y) = \Wiii(x,y) y \cos(x)$ in the disk-slice using the $p$-FEM approach with a single element. Centre: The computed solution to $\Delta u = f$ with zero boundary conditions with $f(x,y) = 1 + \text{erf}(5(1 - 10((x - 0.5)^2 + y^2)))$ in the half-disk. Right: The computed solution to $\Delta u + k^2 \: v \: u = f$ with zero boundary conditions with $f(x,y) = (1-x) \: x \: y \: (1- \half x - y) \: e^x$, $v(x,y) = 1 - (3(x-1)^2 + 5y^2)$ and $k = 100$. in the trapezium.}
%	\centering
%	\label{fig:appendixfigs}
%\end{figure}

As was the case in Section ???, it is possible for our framework in this chapter to be applied to a $p$-finite element method. The ultimate aim here would be to discretise the sphere into spherical cap and spherical band elements, and apply a $p$-finite element method to solve PDEs on the sphere. As a precursor to this, in this section we limit our discretisation to a single spherical cap element, and follow the method of \cite{beuchler2006new} to construct a sparse $p$-finite element method in terms of the operators constructed above, once again with the benefit of ensuring that the resulting discretisation is symmetric. Consider the Dirichlet problem on a surface in 3D $\Omega$:
\begin{align*}
	\begin{cases}
         	- \rho(z)^2 \: \DeltaS u(x,y,z) = \rho(z)^2 \: f(x,y,z) \quad \text{in } \Omega \\
		u = 0 \quad \text{on } \partial \Omega
	\end{cases}
\end{align*}
This has the weak formulation
\begin{align*}
	L(v) := \int_\Omega \rho^2 \: f \: v \: \D \sigma(\bm{x}, z) = \int_\Omega \big(\rho \: \nabla u\big) \cdot \big(\rho \: \nabla v\big) \: \D \sigma(\bm{x}, z) =: a(u,v)
\end{align*}
 for any test function $v \in V := H_0^1(\Omega) = \{v \in H^1(\Omega) \quad | \quad v \rvert_{\partial \Omega} = 0 \}$.

As mentioned, we limit our discretisation to a single element, that is we let $\element = \Omega$ for a spherical cap domain. We can choose our finite dimensional space $V_p = \{v_p \in V \quad | \quad {\rm deg}\,(v_p \rvert_\element) \le p\}$ for some $p \in \N$.

We seek $u_p \in V_p$ s.t.
\begin{align}
	L(v_p) = a(u_p,v_p) \quad \forall \: v_p \in V_p. \label{eqn:FEMweakform}
\end{align}

Define 
\begin{align*}
	\Lambda^{(a)} :=  \ip< \bigscopa_p, \: {\bigscopa_p}^\top >_{\scopa} := \int_\element \bigscopa_p \: {\bigscopa_p}^\top \: \genjacw^{(a,0)} \: \D A = \int_\element \bigW_p^{(a)} \: {\bigscopa_p}^\top \: \D A
\end{align*}
for any $a$. Note that due to orthogonality this is a diagonal matrix. By choosing suitable bases for our FE spaces, we can rewrite (\ref{eqn:FEMweakform}) in matrix-vector form. We can choose a basis for $V_p$ by using the weighted orthogonal polynomials on $\element$ with parameter $a = 1$, the $\bigW_p^{(1)}$ OPs, and we can expand the function $f$ in the $\bigscopi_p$ basis.

Let $\mathbf{u}, \mathbf{v}$ be the coefficient vectors of the expansions of $u_p, v_p \in V_p$ respectively in the $V_p$ basis ($\bigWiii$ OPs). Then,
\begin{align*}
	a(u_p,v_p) &= \int_\element \big(\rho \: \nabla u\big) \cdot \big(\rho \: \nabla v\big) \: \D \sigma(\bm{x}, z) \\
	&= \int_\element \Big( \rhoppphi u_p \: \rhoppphi v_p + \pptheta u_p \: \pptheta v_p \Big) \: \D \sigma(\bm{x}, z) \\
	&= \int_\element \Big( \mathbf{v}^\top {W_\varphi^{(1)}}^\top \bigscopo_p {\bigscopo_p}^\top W_\varphi^{(1)} \mathbf{u} \nonumber \\
					& \quad \quad \quad + \mathbf{v}^\top ({T_W^{(1)\to(0)} D_\theta})^\top \bigscopa_p {\bigscopa_p}^\top T_W^{(1)\to(0)} D_\theta \mathbf{u}  \Big) \: \D \sigma(\bm{x}, z) \\
	&= \mathbf{v}^\top \: \Big( {W_\varphi^{(1)}}^\top \Lambda^{(0)}  W_\varphi^{(1)} + ({T_W^{(1)\to(0)} D_\theta})^\top \Lambda^{(0)} T_W^{(1)\to(0)} D_\theta \Big) \: \mathbf{u}.
\end{align*}
Further, let $\mathbf{f}$ is the coefficient vector for the expansion of the function $f(x,y,z)$ in the $\bigscopi_p$ OP basis and let $P$ be the operator for multiplication by $\rho(z)^2$, that is the operator matrix defined by $P^\top \bigscopi_p(x,y,z) = \rho(z)^2 \: \bigscopi_p(x,y,z)$. Then,
\begin{align*}
	L(v_p) &= \int_\element \: v_p \: \rho^2 \: f \: \D \sigma(\bm{x}, z) \\
	&= \int_\element \: \mathbf{v}^\top \: \bigW_p^{(1)} \: {\bigscopi_p}^\top \: P \: \mathbf{f} \: \D \sigma(\bm{x}, z) \\	
	&= \mathbf{v}^\top \Lambda^{(1)} \: P \: \mathbf{f}.
\end{align*}

Since (\ref{eqn:FEMweakform}) is equivalent to stating that
\begin{align*}
	L(\genjacw^{(1)} \scopnkii) = a(u_p, \genjacw^{(1)} \scopnkii) \quad \forall \: n = 0,\dots,p, \: k = 0,\dots,n, \: i = 0,1
\end{align*}
(i.e. holds for all basis functions of $V_p$) by choosing $v_p$ as each basis function, we can equivalently write the linear system for our finite element problem as:
\begin{align*}
	A\mathbf{u} = \tilde{\mathbf{f}}.
\end{align*}
where the (element) stiffness matrix $A$ is defined by 
\begin{align*}
	A =  {W_\varphi^{(1)}}^\top \Lambda^{(0)}  W_\varphi^{(1)} + ({T_W^{(1)\to(0)} D_\theta})^\top \Lambda^{(0)} T_W^{(1)\to(0)} D_\theta
\end{align*}
and the load vector $\tilde{\mathbf{f}}$ is given by 
\begin{align*}
	\tilde{\mathbf{f}} = \Lambda^{(1)} \: P \: \mathbf{f}.
\end{align*}
Note that since we have sparse operator matrices for partial derivatives and basis-transform, we obtain a symmetric sparse (element) stiffness matrix, as well as a sparse operator matrix for calculating the load vector (rhs).

### 2021-06-18 papers/thesis/submitted/conclusion/summary.tex
Submitted version saved in new dir

In this thesis, we have developed sparse spectral methods for solving partial differential equations (PDEs) on disk-slices and trapeziums in 2D, and spherical caps as a surface in 3D. The work can also be used as a template for developing similar methods on other such similar domains, in particular other subdomains of the sphere.

>>>>

In this thesis, we have developed sparse spectral methods for solving partial differential equations on disk-slices and trapeziums in 2D, and spherical caps as a surface in 3D. The work can also be used as a template for developing similar methods on other such similar domains, in particular other subdomains of the sphere.
